### 2024-10-30 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

Functional diversity is increasingly used alongside taxonomic diversity to describe populations and communities in ecology.
Indeed, functional diversity metrics allow researchers to summarise complex occupancy patterns in space and/or time across communities and/or populations in response to some stressors.
In other words, investigating what, how and why something is changing in an ecosystem by looking at changes of patterns under a certain process through a specific mechanism. 
However, as the diversity of functional diversity metrics and methods increases, it is often not directly clear which metric is more readily appropriate for which question.
We studied the ability of different functional diversity metrics to recover patterns and signals from different processes linked to common assembly mechanisms in community ecology; such as environmental filtering, competitive exclusion, equalising fitness, and facilitation.
Using both simulated data and an empirical dataset affected by more complex and nuanced mechanisms, we tested the effectiveness of different space occupancy metrics to recover the simulated or empirical changes.
We show that different metrics perform better for different tasks, emphasising the importance of not using a one-size-fits-all metric.
Instead, researchers should carefully consider and test whether a particular metric will be effective in capturing a pattern of interest.

@@

Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure \ref{Fig:simulations} and \ref{Tab:mechanisms}).
To apply our framework to a realistic context, we also used an empirical dataset of Hawaiian bird traits and looked at how anthropogenic pressures have shaped the trait space (mechanism; why) based extinction events on Hawaiian islands before and after the year 1500 (process; how) and how different metrics (pattern; what) can lead to different interpretations of the data.
We show that, that with a fixed process and mechanisms approximation, the choice of the statistic to describe the pattern (the disparity or functional diversity metrics) has a great impact on the interpretation of the data with the exact same data being sometimes interpreted in opposite ways.
We suggest caution when summarising patterns in observed data and propose a pipeline and tools for researchers to help understand how their pattern (often stemming from multidimensional data) can vary intuitively or not depending on the process and mechanism of interest.

@@

\begin{table}
\center
\scriptsize
\begin{tabular}{p{0.3\linewidth}|p{0.3\linewidth}|p{0.4\linewidth}}
\textbf{Approximated mechanism (biological)} & \textbf{Used mechanism (algorithm in \texttt{dispRity::reduce.space})} & \textbf{Stressor description}\\
\hline
  Null mechanism & Random removal (``random'') & No overall change in observations' traits or community structure except for the reduction in the number of observations.\\
  Equalising fitness & Size change (``size'') & When observations with trait combinations that are more extreme (i.e. observations located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient. \\
  Facilitation & Density change (``density'') & When observations with closely shared trait combinations (i.e. located close together in the trait space) are more likely to survive the stressor. In other words, this simulates the idea that some observations are more likely to be resilient if they are present in a community of observations (e.g. a community of species) with shared traits rather than the opposite.\\
  Environmental filtering & Position change (``position'') & When an increase in observations' trait similarity happens through some strong abiotic constraints \citep{cornwell2006trait}. For example, some trait combinations become more and more unlikely due to some environmental constraints (i.e. some regions of the trait space become unsuitable).\\
  Competitive exclusion & Evenness change (``evenness'') & When functionally similar observations compete more intensively with one another than with functionally dissimilar observations (``competitive exclusion principle''; \citealt{hardin1960competitive}). At the extreme, it implies that only dissimilar individuals will coexist (``limiting similarity principle''; \citealt{macarthur1967limiting})\\

    \end{tabular}
    \caption{\scriptsize{\textbf{Description of the stressors applied to the simulated data and the mechanism each approximated.} Here we distinguish between the biological mechanism we are trying to simulate and the algorithmic one we used to simulate it. We also provide a more detailed description of the stressor. Note that in biological data, we don't expect any of the mechanisms to act on communities alone (e.g. equalising fitness and facilitation can both act on the distribution of species traits). Nor do we expect their effects to be unidirectional (for example, equalising fitness can happen both by removing the edges or changing the position of a trait distribution). These mechanisms serve as a simplified description of reality for the narrative purpose of this paper.}
}
    \label{Tab:mechanisms}
\end{table}

@@

\textbf{Aspect} & \textbf{Method} & \textbf{Name} & \textbf{Description} & \textbf{\texttt{R} function} & \textbf{Reference}\\
\hline
Richness & Trait Probability Density & Functional richness & Sum of cells where trait probability density $> 0$ & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional richness & Size (volume) of the hypervolume & \texttt{BAT::kernel.alpha} & \cite{mammola2020functional}\\
 & \cr
 & Convex Hull & Functional richness & Volume of the convex hull & \texttt{BAT::hull.alpha} & \cite{cornwell2006trait}\\
 & \cr
 & Dendrogram & Alpha diversity & Total branch length of the functional dendrogram & \texttt{BAT::alpha} & \cite{petchey2002functional,cardoso2015bat}\\
\hline
Divergence & Trait Probability Density & Divergence & Overlap between the TPD function of a single observation and the TPD function of the whole set of observations & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional dispersion  & Average distance between an observation and a sample of random points within and the probabilistic hypervolume & \texttt{BAT::kernel.dispersion} & \cite{mammola2020functional}\\
 & \cr 
 & Dendrogram & Functional dispersion & Average dissimilarity between any two observations in the dendrogram & \texttt{BAT::dispersion} & \cite{cardoso2015bat}\\
 & \cr  
 & Distances & Rao's quadratic entropy & The mean Euclidean distance between every two random pairs of species & \texttt{FD::dbFD()\$RaoQ} & \cite{botta2005rao}\\
\hline
 & Trait Probability Density & Functional evenness & An index of the spread of the trait distribution in an even spread of the distribution (e.g. the height of bins in a histogram) & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional evenness &  Overlap between a hypervolume and an imaginary hypervolume where dimensions are evenly distributed within their possible range & \texttt{BAT::kernel.evenness} & \cite{mammola2020functional}\\
 & \cr 
 & Dendrogram & Functional evenness & Regularity of abundances and distances between observations in the tree. & \texttt{BAT::evenness} & \cite{cardoso2015bat,cardoso2024calculating}\\
 & \cr 
 & Distances & Functional evenness & Regularity of abundances and distances along the minimum spanning tree linking all observations & \texttt{FD::dbFD()\$FEve} & \cite{Laliberte2010FD,villeger2008new}\\
\hline

@@

To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first scaled each metrics by their absolute greatest values to scale them between $0$ and $1$.
We then calculated the difference between each scaled metric score under the stressors of interest (approximating equalisation, facilitation, filtering and competition) and the null stressor for each different stressor level (e.g. $metric_{stressor20\%} - metric_{null20\%}$).
This resulted in each metric being scaled between $–1$ and $1$ where a negative difference can be interpreted as the random removal of elements leading to a higher metric score; a positive difference as the focal removal of elements (i.e. approximating equalisation, facilitation, filtering and competition) leading to a higher metric score; and a difference near $0$ meaning both the null stressor and the focal stressor having a similar effect on the metric score (Figs \ref{Fig:simulations}, \ref{Fig:simulation_results} and \ref{Fig:empirical_results}).
We interpret this last scenario as the metric score performing poorly, i.e. it is unable to clearly distinguish between a random removal of elements and a non random one.
To assess these poor performances, we compared the distributions of the $171$ scores for each metric with each reduction level for each non-null stressor to the null stressor using pairwise t-tests.
The results of these tests are displayed in figure \ref{Fig:simulation_results} (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).

@@

We tested 12 trait space occupancy metrics on simulated and empirical datasets to assess how each metric captures patterns of trait space changes based on stressors approximating ecological and evolutionary mechanisms.
Our results show that different metrics capture different patterns (what) leading to inferring different processes (how) that can be proposed to explain different mechanisms (why).
Therefore, the choice of trait space occupancy metric is essential to accurately describe a pattern of interest.

@@

\section{Conclusion}
Different ecological and evolutionary processes or mechanisms do not always results in different patterns.
Our results based on simulations and empirical data suggest that the same data can be interpreted differently depending on the choice of trait space occupancy metric (\textit{aka} disparity, dissimilarity or functional diversity metrics).
Different metrics are designed to capture different aspects \citep{guillerme2020shifting,mammola2021concepts} but also perform better or worse at their designed task depending on the data, the mechanism or/and process of interest.
Using space occupancy metrics for describing functional diversity is tricky.
This is because the metrics have properties that can be counter intuitive based on the data at hand.
And also because functions of ecosystems or organisms and niches within it are hard to capture/understand/define.
This means that measuring functional diversity is based on both the definition of the metric and the function of interest and thus cannot be served by a one-fits-all metric.
In other words, it is important to choose a disparity/dissimilarity/functional diversity metric depending on the question and the data at hand rather than as a default option or solely based on a previous inspiring publication.

>>>>

Functional diversity is increasingly used alongside taxonomic diversity to describe populations and communities in ecology.
Indeed, functional diversity metrics allow researchers to summarise complex occupancy patterns in space and/or time across communities and/or populations in response to some stressors.
In other words, investigating what, how and why something is changing in an ecosystem by looking at changes of patterns under a certain process through a specific mechanism. 
However, as the diversity of functional diversity metrics and methods increases, it is often not directly clear which metric is more readily appropriate for which question.
We studied the ability of different functional diversity metrics to recover patterns and signals from different processes linked to common assembly mechanisms in community ecology; such as environmental filtering, competitive exclusion, equalising fitness, and facilitation.
Using both simulated data and an empirical dataset affected by more complex and nuanced mechanisms, we tested the effectiveness of different space occupancy metrics to recover the simulated or empirical changes.
We show that different metrics perform differently when trying to capture signal from different approximation of common mechanisms relative to no mechanism at all (null).
For example the competition was harder to disentangle from the null mechanisms compared to facilitation in our simulations.
This emphasises the importance of not using a one-size-fits-all metric.
Instead, researchers should carefully consider and test whether a particular metric will be effective in capturing a pattern of interest.

@@

Through a simulation exercise, we analysed different patterns (what) across different processes (how) approximating different mechanisms (why) of interaction between organisms: equalisation, filtering, facilitation and competition (Figure \ref{Fig:simulations} and \ref{Tab:mechanisms}).
Our goal is to the relative performance of different metrics to capture the patterns of different mechanisms by comparing the score of various metrics under specifically approximated mechanisms and the absence of any specific mechanism (null mechanism).
We expect that our ability to capture a pattern does not only depends on the choice of metric (what) but also on the process and mechanism at hand (how and why).
To apply our framework to a realistic context, we also used an empirical dataset of Hawaiian bird traits and looked at how anthropogenic pressures have shaped the trait space (mechanism; why) based extinction events on Hawaiian islands before and after the year 1500 (process; how) and how different metrics (pattern; what) can lead to different interpretations of the data.
We show that, that with a fixed process and mechanisms approximation, the choice of the statistic to describe the pattern (the disparity or functional diversity metrics) has a great impact on the interpretation of the data with the exact same data being sometimes interpreted in opposite ways.

@@

\begin{table}
\center
\scriptsize
\begin{tabular}{p{0.2\linewidth}|p{0.2\linewidth}|p{0.4\linewidth}|p{0.2\linewidth}}
\textbf{Approximated mechanism (biological)} & \textbf{Used mechanism (algorithm in \texttt{dispRity::reduce.space})} & \textbf{Stressor description} & \textbf{Type of metric expected to recover the mechanism}\\
\hline
  Null mechanism & Random removal (``random'') & No overall change in observations' traits or community structure except for the reduction in the number of observations. & Metrics sensitive to the amount of data \\
  Equalising fitness & Size change (``size'') & When observations with trait combinations that are more extreme (i.e. observations located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient. & Metrics capturing changes in richness (but not divergence and regularity) \\
  Facilitation & Density change (``density'') & When observations with closely shared trait combinations (i.e. located close together in the trait space) are more likely to survive the stressor. In other words, this simulates the idea that some observations are more likely to be resilient if they are present in a community of observations (e.g. a community of species) with shared traits rather than the opposite. & Metrics capturing changes in regularity (but not richness and regularity) \\
  Environmental filtering & Position change (``position'') & When an increase in observations' trait similarity happens through some strong abiotic constraints \citep{cornwell2006trait}. For example, some trait combinations become more and more unlikely due to some environmental constraints (i.e. some regions of the trait space become unsuitable). & Metrics capturing changes in position (and richness) \\\\
  Competitive exclusion & Evenness change (``evenness'') & When functionally similar observations compete more intensively with one another than with functionally dissimilar observations (``competitive exclusion principle''; \citealt{hardin1960competitive}). At the extreme, it implies that only dissimilar individuals will coexist (``limiting similarity principle''; \citealt{macarthur1967limiting}). & Metrics capturing changes in divergence (but not richness and regularity) \\\\

    \end{tabular}
    \caption{\scriptsize{\textbf{Description of the stressors applied to the simulated data and the mechanism each approximated.} Here we distinguish between the biological mechanism we are trying to simulate and the algorithmic one we used to simulate it. We also provide a more detailed description of the stressor. Note that in biological data, we don't expect any of the mechanisms to act on communities alone (e.g. equalising fitness and facilitation can both act on the distribution of species traits). Nor do we expect their effects to be unidirectional (for example, equalising fitness can happen both by removing the edges or changing the position of a trait distribution), thus, the type of metric expected to recover the mechanism can be variable. These mechanisms serve as a simplified description of reality for the narrative purpose of this paper.}
}
    \label{Tab:mechanisms}
\end{table}

@@

\textbf{Aspect} & \textbf{Method} & \textbf{Name} & \textbf{Description} & \textbf{\texttt{R} function} & \textbf{Reference}\\
\hline
Richness & Trait Probability Density & Functional richness & Sum of cells where trait probability density $> 0$ & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional richness & Size (volume) of the hypervolume & \texttt{BAT::kernel.alpha} & \cite{mammola2020functional}\\
 & \cr
 & Convex Hull & Functional richness & Volume of the convex hull & \texttt{BAT::hull.alpha} & \cite{cornwell2006trait}\\
 & \cr
 & Dendrogram & Alpha diversity & Total branch length of the functional dendrogram & \texttt{BAT::alpha} & \cite{petchey2002functional,cardoso2015bat}\\
\hline
Divergence & Trait Probability Density & Divergence & Overlap between the TPD function of a single observation and the TPD function of the whole set of observations & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional dispersion  & Average distance between an observation and a sample of random points within and the probabilistic hypervolume & \texttt{BAT::kernel.dispersion} & \cite{mammola2020functional}\\
 & \cr 
 & Dendrogram & Functional dispersion & Average dissimilarity between any two observations in the dendrogram & \texttt{BAT::dispersion} & \cite{cardoso2015bat}\\
 & \cr  
 & Distances & Rao's quadratic entropy & The mean Euclidean distance between every two random pairs of species & \texttt{FD::dbFD()\$RaoQ} & \cite{botta2005rao}\\
\hline
Regularity & Trait Probability Density & Functional evenness & An index of the spread of the trait distribution in an even spread of the distribution (e.g. the height of bins in a histogram) & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional evenness &  Overlap between a hypervolume and an imaginary hypervolume where dimensions are evenly distributed within their possible range & \texttt{BAT::kernel.evenness} & \cite{mammola2020functional}\\
 & \cr 
 & Dendrogram & Functional evenness & Regularity of abundances and distances between observations in the tree. & \texttt{BAT::evenness} & \cite{cardoso2015bat,cardoso2024calculating}\\
 & \cr 
 & Distances & Functional evenness & Regularity of abundances and distances along the minimum spanning tree linking all observations & \texttt{FD::dbFD()\$FEve} & \cite{Laliberte2010FD,villeger2008new}\\
\hline

@@

To understand the ability of each metric to capture a mechanism of interest, we compared their scores relatively to the same metric scores of a null mechanism.
To ease interpretation of the results across a range of metrics with different orders of magnitudes of outputs and different random simulations, we first scaled each metrics by their absolute greatest values to scale them between $0$ and $1$.
We then calculated the difference between each scaled metric score under the stressors of interest (approximating equalisation, facilitation, filtering and competition) and the null stressor for each different stressor level (e.g. $metric_{stressor20\%} - metric_{null20\%}$).
This resulted in each metric being scaled between $–1$ and $1$ where a negative difference can be interpreted as the random removal of elements leading to a higher metric score; a positive difference as the focal removal of elements (i.e. approximating equalisation, facilitation, filtering and competition) leading to a higher metric score; and a difference near $0$ meaning both the null stressor and the focal stressor having a similar effect on the metric score (Figs \ref{Fig:simulations}, \ref{Fig:simulation_results} and \ref{Fig:empirical_results}).
We interpret this last scenario as the metric score performing poorly, i.e. it is unable to clearly distinguish between a random removal of elements and a non random one.
To assess these poor performances, we compared the distributions of the $171$ scores for each metric with each reduction level for each non-null stressor to the null stressor using pairwise t-tests.
The results of these tests are displayed in figure \ref{Fig:simulation_results} (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).

@@

We tested 12 trait space occupancy metrics on simulated and empirical datasets to assess how each metric captures patterns of trait space changes based on stressors approximating ecological and evolutionary mechanisms.
Our results show that different metrics capture different patterns (what) leading to inferring different processes (how) that can be proposed to explain different mechanisms (why).
Therefore, the choice of trait space occupancy metric is essential to accurately describe a pattern of interest.
We found that different metrics can capture changes in pattern relative to the null mechanisms in different ways depending on the approximated mechanism of interest.
For example, for the approximated equalising mechanism, most metrics capture a pattern clearly smaller than the one from the null mechanism (i.e. randomly removing elements has more effects on the metric than removing them following the mechanism).
However, some metrics (e.g. the functional evenness based on the hypervolume) capture a pattern clearly greater than the one from the null mechanism.
This means that changes in metric values can be counter intuitive and should always been interpreted in the context of the metric, data and question at hand.

For example, say one is interested in looking at changes in functional diversity using a divergence metric using the probability density method (e.g. \texttt{TPD::REND}).
They can then apply this metric to their trait space and get two relative metric scores of, says, 0.7 and 0.65 (pattern) for two levels of extinction (process).
The interpretation of this score as a change in functional diversity will depend on the worker's assumption of which mechanism affected the trait space.
In our simulations, if the mechanism approximated was the equalising one, the pattern can clearly be interpreted as a change in functional diversity, but if the mechanism is a facilitation one, the pattern can be meaningless (i.e. not clearly distinguishable from a null mechanism - Fig. \ref{Fig:simulation_results} shows this example by contrasting row 5, columns 1 and 2).
In this made up example, the procedure is done in reverse order where the the pattern is specified first and then the process and the mechanism (``What process and mechanism are needed to interpret pattern X?'').
In Box 2 we suggest an approach that specifies the process and the mechanism first (``how do I interpret pattern X under a specific process and mechanism''; see Box 2).

Regarding the different approximated mechanisms, it is interesting to note that our different approximations lead to different general behaviours of the metrics: for example the range of variance of metrics behaviour is greater for the equalising mechanism and smaller for the competition mechanism (albeit nearly always distinguishable from the null mechanism).
This differences in metrics variance per mechanisms suggest that some mechanisms leads to more subttle changes or/and that our approximation of these mechanisms can be variable (see caveats below).

@@

\section{Conclusion}
Different ecological and evolutionary processes or mechanisms do not always results in different patterns.
Our results based on simulations and empirical data suggest that the same data can be interpreted differently depending on the choice of trait space occupancy metric (\textit{aka} disparity, dissimilarity or functional diversity metrics).
Different metrics are designed to capture different aspects \citep{guillerme2020shifting,mammola2021concepts} but also perform better or worse at their designed task depending on the data, the mechanism or/and process of interest.
Using space occupancy metrics for describing functional diversity is tricky.
This is because the metrics have properties that can be counter intuitive based on the data at hand.
And also because functions of ecosystems or organisms and niches within it are hard to capture/understand/define.
This means that measuring functional diversity is based on both the definition of the metric and the function of interest and thus cannot be served by a one-fits-all metric.
In other words, it is important to choose a disparity/dissimilarity/functional diversity metric depending on the question and the data at hand rather than as a default option or solely based on a previous inspiring publication.
We suggest caution when summarising patterns in observed data and propose a pipeline and tools for researchers to help understand how their pattern (often stemming from multidimensional data) can vary intuitively or not depending on the process and mechanism of interest.
